### 2024-09-11 Article/main.tex
Lemma2.2.1, Theroem2.2.2, without proof

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{prop}[theorem]{Proposition}

@@

\section{Functors, Natural transformations and Beyond}\label{sect:funct_nats}
\todo{Toglierei il "beyond" DC}
A functor is a structure preserving map between categories. 
\todo{E cambierei anche l'intro a questa sezione DC}
\subsection{Functors}
\begin{definition}[Functor]
    Let $\cat{C}$ and $\cat{D}$ be categories. A \emph{functor} $F:\cat{C \rightarrow D}$ is a map taking each object of $A \in \Ob(\cat{C})$ to an object $F(A) \in \Ob(\cat{D})$ and each arrow $f:A\rightarrow B$ of $\cat C$ to a arrow $F(f): F(A) \rightarrow F(B)$ of $\cat D$, such that, for all objects $A \in \Ob(\cat C)$ and composable arrows $f$ and $g$ of $\cat C$:
    \begin{itemize}
        \item $F(id_{A}) = id_{F(A)}$;
        \item $F(g \circ f) = F(g) \circ F(f)$.
    \end{itemize}

@@

\begin{obs}\label{obs:funct_creat_lim_then_lift}%[~\cite{Adámek_Herrlich_Strecker_2009}]
    It holds that if a functor creates limits, then lifts uniquely limits.
\end{obs}

@@

@@

Previous examples show that the category $\mathbf{EqGrph}$ has both (c)equalizer and finite (co)products, so, by Theorem~\ref{th:limit}, we can conclude that $\mathbf{EqGrph}$ has finite limits and colimits.

@@

Considering the category of graphs (as defined in Definition~\ref{def:cat_of_graph}), it is possible to show that 
$$\mathbf{Graph} \cong [E \mathrel{\mathop{\rightrightarrows}^{s}_{t}} V , \Set]$$
where an object $G$ is a functor, having $G(V)$ as the set of vertices, $G(E)$ as the set of edges, $G(s)$ and $G(t)$ as the source and target functions. A morphisms $\eta: G \dot\rightarrow H$ is a natural trasformation, and the commutativity of the diagram below is given by definition of graph morphisms (Definition~\ref{def:graph_hom})
$$
    \begin{tikzcd}
        G(E) \arrow[r, "{\eta_E}"] \arrow[d, "G(s)"swap] & H(E) \arrow[d, "H(s)"] \\
        G(V) \arrow[r, "{\eta_V}"swap] & H(V)
    \end{tikzcd}
$$
(and the same for $t$).

Example~\ref{ex:in_term_in_graph} and following, in which limits and colimits are computed component-wise, are a concrete instance of Observation~\ref{obs:limits_in_presh}.

More complex objects are graphs with equivalence. We can think of $\mathbf{EqGrph}$ as a subcategory of the category $[V \rightrightarrows E \rightarrow Q, \Set]$, in which each object $\mathcal{G}$ is a functor mapping the arrow $q: E \rightarrow Q$ onto an epimorphism (i.e., a surjective function). The component $\mathcal{G}(q)$ here is intended to be the function discussed in Remark~\ref{rem:eq_as_surj}.

>>>>

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{cor}[theorem]{Corollary}

@@

\section{Functors and Natural Transformations}\label{sect:funct_nats}

A functor is a structure preserving map between categories. 
\todo{E cambierei anche l'intro a questa sezione DC}

\subsection{Functors}
\begin{definition}[Functor]
    Let $\cat{C}$ and $\cat{D}$ be categories. A \emph{functor} $F:\cat{C \rightarrow D}$ is a map taking each object of $A \in \Ob(\cat{C})$ to an object $F(A) \in \Ob(\cat{D})$ and each arrow $f:A\rightarrow B$ of $\cat C$ to a arrow $F(f): F(A) \rightarrow F(B)$ of $\cat D$, such that, for all objects $A \in \Ob(\cat C)$ and composable arrows $f$ and $g$ of $\cat C$:
    \begin{itemize}
        \item $F(id_{A}) = id_{F(A)}$;
        \item $F(g \circ f) = F(g) \circ F(f)$.
    \end{itemize}

@@

\begin{obs}\label{obs:funct_creat_lim_then_lift}
    It holds that if a functor creates limits, then lifts uniquely limits~\cite{Adamek_Herrlich_Strecker_2009}.
\end{obs}

@@

Graphs with equivalences are alternative representation of \emph{quotient graphs} over an equivalence relation. The graph with equivalence $\mathcal{G} = (G, \sim_G)$ is another representation of the graph $G/_{\sim_G}$.

\begin{definition}\label{def:quot_func}[Quotient Functor]
    The \emph{quotient functor} $Q: \mathbf{EqGrph} \rightarrow \mathbf{Graph}$ is defined as follows:
    \begin{itemize}
        \item on objects, given $\mathcal{G}=(G, \sim_{G})$ as $Q(\mathcal{G}) = G/_{\sim_G} = (V/_{\sim_G}, E, s', t')$, where $s'(e) = [v]_{\sim_G}$ if $s(e) = v$, and $t'(e) = [v]_{\sim_G}$ if $t(e) = v$;
        \item on morphisms, given $h = (h_V, h_E): \mathcal{G \rightarrow H}$, as $Q(h_V)([v]_{\sim_G}) = [h_V(v)]_{\sim_H}$, and $Q(h_E) = h_E$.
    \end{itemize}
\end{definition}

@@

Previous examples show that the category $\mathbf{EqGrph}$ has both (co)equalizers and finite (co)products, so, by Theorem~\ref{th:limit}, we can conclude that $\mathbf{EqGrph}$ has finite limits and colimits.

An important characterization in the category of graphs with equivalences is the following.

\begin{prop}\label{prop:reg_mono_in_EG_are_mono_in_graph}
    In $\mathbf{EqGrph}$, a monomorphism is regular if it reflects equivalences. That is, if $h: \mathcal{G \rightarrow H}$ is mono and such that $h_V (v_1) \sim_H h_V(v_2) \Rightarrow v_1 \sim_G v_2$, then $h$ is regular mono.
\end{prop}

@@

Considering the category of graphs (as defined in Definition~\ref{def:cat_of_graph}), it is possible to show that 
$$\mathbf{Graph} \cong [E \mathrel{\mathop{\rightrightarrows}^{s}_{t}} V , \Set]$$
where an object $G$ is a functor, having $G(V)$ as the set of vertices, $G(E)$ as the set of edges, $G(s)$ and $G(t)$ as the source and target functions. A morphisms $\eta: G \dot\rightarrow H$ is a natural transformation, and the commutativity of the diagram below is given by definition of graph morphisms (Definition~\ref{def:graph_hom})
$$
    \begin{tikzcd}
        G(E) \arrow[r, "{\eta_E}"] \arrow[d, "G(s)"swap] & H(E) \arrow[d, "H(s)"] \\
        G(V) \arrow[r, "{\eta_V}"swap] & H(V)
    \end{tikzcd}
$$
(and the same for $t$).

Example~\ref{ex:in_term_in_graph} and following, in which limits and colimits are computed pointwise, are a concrete instance of Observation~\ref{obs:limits_in_presh}.

More complex objects are graphs with equivalence. We can think of $\mathbf{EqGrph}$ as a subcategory of the category $[V \rightrightarrows E \rightarrow Q, \Set]$, in which each object $\mathcal{G}$ is a functor mapping the arrow $q: E \rightarrow Q$ onto an epimorphism (i.e., a surjective function). The component $\mathcal{G}(q)$ here is intended to be the function discussed in Remark~\ref{rem:eq_as_surj}. 
In the following, we will refer with $\mathbf{Graph}$ and $\mathbf{EqGrph}$ to the category of presheaves we have just introduced.

Regular monos in $\mathbf{EqGrph}$ are then natural transformations $\eta: \mathcal{G \dot\rightarrow H}$ such that $\eta_Q$ is mono (Proposition~\ref{prop:reg_mono_in_EG_are_mono_in_graph}).

\begin{lemma}
    $\mathbf{EqGrph}$ is stable under $\Reg(\mathbf{EqGrph})$-pushouts and $\Reg({\mathbf{EqGrph}})$-pullbacks.
\end{lemma}

\begin{theorem}
    $EqGrph$ is $\Reg(\mathbf{EqGrph})$-adhesive.
\end{theorem}
